---
title: "p8105_hw2_zl3543"
author: "Zihan Lin"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Problem 1
# Load necessary libraries
library(tidyverse)

# Read the CSV data file
subway_data <- read_csv("/Users/suwa/Desktop/p8105_hw2_zl3543/data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv")

# Retain relevant columns and convert 'entry' to a logical variable
cleaned_data <- subway_data %>%
  select(line = `Line`, 
         station_name = `Station Name`, 
         station_latitude = `Station Latitude`, 
         station_longitude = `Station Longitude`, 
         routes = `Route1`:`Route11`, 
         entry = `Entry`, 
         vending = `Vending`, 
         entrance_type = `Entrance Type`, 
         ada_compliant = `ADA`) %>%
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE))

# Check the dimensions of the cleaned dataset
dim(cleaned_data)

# The dataset contains information about entrances and exits for each subway station in NYC. Key variables include 'line', 'station_name '(station name)', 'station_latitude', and 'station_longitude', 'routes', 'entry', 'vending', 'entrance_type', and 'ada_compliant'. For data cleaning, we retained only the relevant columns, renamed them for clarity, and converted the entry column from a character "YES"/"NO" to a logical 'TRUE'/'FALSE'. The resulting cleaned dataset has dimensions of [1868 rows x 19 columns], where row is the number of entries and column represents the selected variables. This dataset is tidy as each variable is in its own column, each observation is a row, and each cell contains a single value.

# Count distinct stations
distinct_stations <- cleaned_data %>%
  distinct(line, station_name)

num_distinct_stations <- nrow(distinct_stations)
num_distinct_stations

# Include 'ada_compliant' in the distinct selection
distinct_stations <- cleaned_data %>%
  distinct(line, station_name, ada_compliant)

# Count ADA compliant stations
num_ada_compliant_stations <- distinct_stations %>%
  filter(ada_compliant == TRUE) %>%
  nrow()

num_ada_compliant_stations

# Calculate the proportion of entrances/exits without vending that allow entry
no_vending_entries <- cleaned_data %>%
  filter(vending == "NO") %>%
  summarise(proportion = mean(entry, na.rm = TRUE))

no_vending_entries

# Convert all route columns to character to avoid type inconsistency
cleaned_data <- cleaned_data %>%
  mutate(across(starts_with("route"), as.character))

# Gather route columns into long format
route_data <- cleaned_data %>%
  pivot_longer(cols = starts_with("route"), 
               names_to = "route_number", 
               values_to = "route_name") %>%
  filter(!is.na(route_name))

# Display the first few rows to check the result
head(route_data)

# Count distinct stations serving the A train
stations_serving_A <- route_data %>%
  filter(route_name == "A") %>%
  distinct(line, station_name) 

num_stations_A <- nrow(stations_serving_A)
num_stations_A

# Find the number of ADA compliant stations serving the A train
ada_compliant_A <- route_data %>%
  filter(route_name == "A", ada_compliant == "TRUE") %>%
  distinct(line, station_name) %>%
  nrow()

ada_compliant_A

```

```{r}
# Problem 2
# Load necessary libraries
library(tidyverse)
library(readxl)
library(janitor)


# Read the Mr.Trash Wheel sheet, specifying the sheet and skipping non-data rows
mr_trash_wheel <- read_excel("/Users/suwa/Desktop/p8105_hw2_zl3543/data/202309 Trash Wheel Collection Data.xlsx", 
                             sheet = "Mr. Trash Wheel",
                             skip = 1) %>%
  # Rename columns to have reasonable variable names
  clean_names() %>%
  # Remove rows with missing dumpster-specific data
  filter(!is.na(dumpster)) %>%
  # Round the number of sports balls and convert to integer
  mutate(sports_balls = as.integer(round(sports_balls, 0)))

# Read the Professor Trash Wheel sheet
prof_trash_wheel <- read_excel("/Users/suwa/Desktop/p8105_hw2_zl3543/data/202309 Trash Wheel Collection Data.xlsx", 
                               sheet = "Professor Trash Wheel", 
                               skip = 1) %>%
  clean_names() %>%
  filter(!is.na(dumpster)) %>%
  mutate(sports_balls = NA_integer_,  # Add sports_balls column as NA
         trash_wheel = "Professor Trash Wheel")

# Read and clean the Gwynnda Trash Wheel data
gwynnda_trash_wheel <- read_excel("/Users/suwa/Desktop/p8105_hw2_zl3543/data/202309 Trash Wheel Collection Data.xlsx", 
                                  sheet = "Gwynnda Trash Wheel", 
                                  skip = 1) %>%
  clean_names() %>%
  filter(!is.na(dumpster)) %>%
  mutate(sports_balls = NA_integer_,  # Add sports_balls column as NA
         trash_wheel = "Gwynnda Trash Wheel")

# Convert 'year' to numeric in all datasets
mr_trash_wheel <- mr_trash_wheel %>%
  mutate(year = as.numeric(year))

prof_trash_wheel <- prof_trash_wheel %>%
  mutate(year = as.numeric(year))

gwynnda_trash_wheel <- gwynnda_trash_wheel %>%
  mutate(year = as.numeric(year))

# Combine all datasets
combined_trash_wheel <- bind_rows(mr_trash_wheel, prof_trash_wheel, gwynnda_trash_wheel)

# Display the dimensions and the first few rows to check the result
dim(combined_trash_wheel)
head(combined_trash_wheel)

```

The combined dataset contains information on trash collection from three different trash wheels: Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda. There are a total of `r nrow(combined_trash_wheel)` observations and `r ncol(combined_trash_wheel)` variables in the resulting dataset. Key variables include `dumpster`, which identifies the dumpster number, `date` indicating when trash was collected, `weight_tons` representing the weight of trash collected in tons, `plastic_bottles` showing the number of plastic bottles collected, and `sports_balls` counting the number of sports balls recovered. The total weight of trash collected by Professor Trash Wheel is `r sum(combined_trash_wheel %>% filter(trash_wheel == "Professor Trash Wheel") %>% pull(weight_tons), na.rm = TRUE)` tons. Additionally, Gwynnda collected `r sum(combined_trash_wheel %>% filter(trash_wheel == "Gwynnda", lubridate::year(date) == 2022, lubridate::month(date) == 6) %>% pull(cigarette_butts), na.rm = TRUE)` cigarette butts in June 2022.

